---
title: "Tidy Tuesday 3: Registered Nurses"
description: |
  Tidy Tuesday from the week of October 5 using data on registered nurses and the ggmap package.
author:
  - name: Mia Rothberg
    url: {}
date: 10-05-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error=TRUE, 
                      message=FALSE, 
                      warning=FALSE)
```

```{r libraries}
library(tidyverse)     # for graphing and data cleaning
library(ggthemes)      # for even more plotting themes
library(scales)
library(plotly)
library(ggmap)
library(tmaptools)
library(maps)
```

```{r data}
# Read in the data for the week
nurses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-05/nurses.csv')

population <- read.csv("tidy-tues-nst-est2020.csv")

```

```{r}
population <- population %>%
  filter(STATE > 0) %>% 
  filter(STATE < 57) %>% 
  select(NAME, POPESTIMATE2010, POPESTIMATE2020)
```

```{r}
nurses_joined <- nurses %>% 
  left_join(population, by = c("State" = "NAME"))
```

```{r}
nurses_2019_2020 <- nurses_joined %>% 
  filter(Year == 2010 | Year == 2020) %>% 
  mutate(std_registered_per_1000 = ifelse(Year == 2010, `Total Employed RN`/POPESTIMATE2010 * 1000, `Total Employed RN`/POPESTIMATE2020 * 1000)) %>% 
  mutate(`Registered Nurses per 1000 People` = std_registered_per_1000) %>% 
  mutate(state_lower = tolower(State))
```

```{r}
states_map <- map_data("state")

nurses_map <- nurses_2019_2020 %>% 
  filter(Year == 2020) %>% 
  ggplot(aes(fill = std_registered_per_1000, label = State, label2 = `Registered Nurses per 1000 People`, label3 = `Annual Salary Median`)) +
  geom_map(aes(map_id = state_lower), color="gray", size=0.3, map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() +
  scale_fill_gradient(low="white", high = "navy") +
  labs(fill = "Registered Nurses per \n1000 People in 2020")

#ggplotly(nurses_map, tooltip = c("label", "label2", "label3"))



plotly_map <- plotly::ggplotly(nurses_map, tooltip = c("label", "label2", "label3"))

htmlwidgets::saveWidget(plotly_map, "nurses_plotly_map.html")
```

```{r, results='asis'}
print( htmltools::tags$iframe(src="nurses_plotly_map.html", width=600, height=480) )
```


> For this Tidy Tuesday, I looked at the geospatial relationship of number of registered nurses per 1000 people in a state. I decided to add animation via `ggplotly` to display the state name, the exact number of nurses, and the median annual salary of nurses in that state. I was surprised to see that South Dakota had the most nurses per population, perhaps because it has the lowest population.